Title

### 2024-06-26 paper/main.tex
Updates

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

@@

Many questions in science center around understanding the cause and effect
relationships between variables, as it can reveal the underlying mechanisms for
the observed phenomena and inform effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to find the underlying causal structure
among random variables from a given dataset. The problem of causal discovery
has been studied in both Directed Acyclic Graphs (DAGs) and Structural Equation
Models (SEM) literatures with different approaches. In the DAG literature, the
focus has been on developing automated algorithms that can learn the causal
structure from a given observational or interventional dataset, whereas in the
SEM literature the focus has on developing tools to help researchers construct 
models manually or modify manually constructed models.

@@

In the DAG literature, many automated algorithms have been developed for causal
discovery such as constraint based methods like PC \citep{Spirtes2001} and FCI
\citep{Spirtes2000}, score-based methods like Hill-Climb Search, Greedy
Equivalence Search \citep{Chickering2002}, and optimization based methods.
However, the adaption of these algorithms in applied fields remain limited
\citep{Tennant2020}. In applied fields, researchers typically prefer to
construct these DAGs manually based on their domain knowledge. We hypothesize
that this preference towards manual DAG construction stems from the following
difficulties in applying causal discovery methods:

\begin{enumerate}
	\item \textbf{Difficult to trust: } Even though most of these algorithms
		have asymptotic theory proving them to be consistent, their
		finite sample properties is not well understood. In real
		setting with finite samples, these algorithms can make obvious
		mistakes. As a result of this, researchers have to rely on
		post-hoc modification of the output of these algorithms
		using either their domain knowledge or assisted by model testing
		methods.
	\item \textbf{Algorithms output Markov Equivalence Class (MEC):} Given an 
		observational dataset, multiple causal structures can be
		faithful to it. As a result, using only observational datasets,
		these algorithms can only get to the MEC such as Completed
		Partially Directed Acyclic Graph (CPDAG) or a Partial Ancestral
		Graph (PAG) where not all the edges are directed. However, as
		most of the methods developed for downstream tasks such as
		identification or casual effect estimation assume the knowledge
		of a DAG. To orient these MECs to a DAG, researchers have to
		either rely on their domain expertise, or make further
		assumption to utilize a pairwise orientation method.
	\item \textbf{Difficult to choose appropriate algorithm and
		hyperparameters:} All the algorithms have the same goal with
		all of them having similar asymptotic properties. However, the
		finite sample properties of these methods vary depending on the
		choices that we make. For example, the choice of the
		conditional independence (CI) test being used for
		constraint-based method depends on the dataset, similarly for
		the scoring method for score-based methods. With no obvious way
		to test the impact/performance of these choices on a given
		dataset, it becomes quite hard to chose the optimal algorithm
		for a given dataset.
\end{enumerate}

On the other hand, in the field of Structural Equation Models (SEMs), a lot
more focus has been put on methods to help researchers draw models manually
along with integrating their expertise. Researchers typically come up with an
initial model based on their domain knowledge, and then use various tools to
guide them in making modifications to this initial model. This process of
constructing models is commonly known as \emph{Specification Search}. A
commonly used method for suggesting potential modifications to the SEM is
\emph{modification indices}. Given a specified model, modification indices
tries to find potential modifications to the model that improve the model's fit
to the data best. Based on the suggestions from this method along with their
domain expertise, the researcher can then choose the most optimal modification
for their model.

\todo[inline]{Other methods: Modification indices and Lagrange Multiplier tests to
selectively add edges And using Wald test or z statistics (also known as
critical ratios) for removing edges.}

As it is evident that researchers also draw DAGs by hand and even if they use
an algorithm, the need to make some post-hoc adjustments to the model, in this
paper we present a tool to assist researchers in making these adjustments
somewhat inspired by modification indices. In the case of DAGs, global testing
methodology does exist such as Shipley test, that can combine local structure
tests into a single p-value test. However, this does not give us a way to make
modifications to our model. A possible way to do this by checking which of
these local tests are violated. However, this approach can quickly become
diffult as the size of the network increases. For example, the popular alarm
network implies $x$ condition independence statements.

@@

@@

Our main contributions in this papers are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this association measure, we present a procedure to suggest
		potential new edges, or edge removals in the model. The user
		gets to decide the direction of the new edge based on their
		expertise.
	\item Lastly, we present a web-browser based tool to allow researchers
		to apply this method to their own datasets.
\end{enumerate}

@@

				\draw[->]  (age) to[bend right=20] (ed);
				\draw[->]  (sex) to (ed);
				\draw[->]  (age) to (hrpw);
				\draw[->]  (ed) to (hrpw);
				\draw[->]  (sex) to (hrpw);
				\draw[->]  (ed) to (income);
				\draw[->]  (hrpw) to (income);
				\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
				\draw[->]  (sex) to (income);
				\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
				\draw[->]  (sex) to (occ);	
			\end{scope}
			\begin{scope}[xshift=-3cm]
				\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
				\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
					\begin{minipage}{.2\textwidth}
						\begin{equation*}
							\begin{split}
								\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
								\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
								\textnormal{Age} &\ci \textnormal{Sex} \\
								\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\end{split}
						\end{equation*}
					\end{minipage}
					};
			\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Show a full example of testing using this model}}
\end{figure}

@@

\section{Background}
\label{sec:background}
We represent uni-dimensional variables as $ X $ and multi-dimensional variables
as $ \bm{Z} $. All the variables are considered to be any mix of continuous,
ordinal, or categorical unless specified explicitly.

We consider the case of estimating association between $ X $ and $ Y $ is the
presence of a set of conditioning variables $ \bm{Z} = Z_1, \cdots, Z_k $,
potentially being an empty set $ \bm{Z} = \emptyset $. We use $ \rvert \bm{Z}
\rvert $ to denote the cardinality of the set of variables $ \bm{Z} $. We
consider various scenarios where the variables $ X $, $ Y $ and $ \bm{Z} $
could be either discrete, ordinal, or continuous. With mixed data, we mean an
arbitrary mix of variable types among $ X $, $ Y $, and $ \bm{Z} $.

We consider the problem of estimating the partial (or conditional) association
between $ X $ and $ Y $ in the presence of the conditioning set $ Z $. When $ Z
= \emptyset $, this is equivalent to the marginal association between $ X $ and
$ Y $.

@@

In this section, we develop a novel (conditional) measure of association for
mixed data by extending the idea of partial correlation to mixed data. We do
this by combining a mixed-data residualization method with a multivariate
measure of association. Given two variables $ X $ and $ Y $, and a set of
conditional variables $ \bm{Z} $, we are interested in computing an estimate of
the conditional association, $ \rho(X, Y; Z) $. If $ \bm{Z} = \emptyset $, we
use a constant value as $ Z $. Firstly, we compute the residuals for $ x $ and
$ y $ using the $ \bm{z} $ as the predictor variables. We use the mixed data
residualization approach from \citet{Ankan2023}. We consider each variable type
separately and repeat the following steps for both variables $ X $ and $ Y $:

@@

@@

The idea behind using the measure of association for modifying DAGs is inspired
by modification indices in SEM.

Theorem - If we have access to an oracle that can give all correct results, 
the algorithm above is consistent.

>>>>

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}

@@

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific questions as it reveals the underlying mechanisms of
the observed phenomena and informs effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to uncover the underlying causal structure
among random variables from a given dataset. The problem of causal discovery
has been explored in both the Directed Acyclic Graphs (DAGs) and Structural
Equation Models (SEM) literatures, each with different approaches. In the DAG
literature, the focus has been on developing automated algorithms that can
learn the causal structure from observational or interventional dataset. In
contrast, the SEM literature emphasizes tools that help researchers manually
construct or modify modify.

@@

In the DAG literature, many automated algorithms have been developed for causal
discovery such as constraint based methods like PC \citep{Spirtes2001} and FCI
\citep{Spirtes2000}, score-based methods like Hill-Climb Search, Greedy
Equivalence Search \citep{Chickering2002}, and optimization based methods.
However, the adaption of these algorithms in applied fields remain limited
\citep{Tennant2020}. Researchers in applied fields typically prefer to
construct DAGs manually based on their domain knowledge. We hypothesize that
this preference towards manual DAG construction stems from the following
challenges in applying causal discovery algorithms:

\begin{enumerate}
	\item \textbf{Trustworthiness: } Although most algorithms have
		asymptotic theory proving them to be consistent, their finite
		sample properties are not well understood. In real-world
		settings with finite samples, these algorithms can make obvious
		mistakes, for example, the constraint-based algorithms are
		biased towards sparse graphs \citep{Malinsky2024}. Consequently,
		researchers often have rely on post-hoc modifications based on 
		their domain knowledge or model testing methods.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple causal
		structures can be faithful to a given observational dataset.
		Thus, using only observational data, these algorithms can only
		recover the MECs that can contain unoriented edges. However,
		most methods for downstream tasks, such as identification or
		causal effect estimation, assume knowledge of a fully oriented
		DAG. As a results, researchers have to rely on their expertise
		to orient these edges or make further assumptions to make use
		of pairwise orientation techniques.
	\item \textbf{Algorithm and hyperparameter selection:} Each algorithm
		has the same goal but they can learn very different models
		depending on the choice of algorithm and their hyperparameters.
		For example, the choice of the conditional independence (CI)
		test being used for constraint-based method depends on the
		dataset, similarly for the scoring method for score-based
		methods. With no obvious way to test the impact/performance of
		these choices on a given dataset, it becomes quite hard to
		chose the optimal algorithm for a given dataset.
\end{enumerate}

As a result of these challenges, no matter if the researcher is manually
constructing the DAG or using an algorithm, they need to utilize their domain
expertise to construct and modify models. One potential way to through
conditional independence (CI) based testing. As each missing edge in the DAG
implies a CI statement, that can be tested in our data \ref{fig:ci_testing}.
This however has two main drawbacks: 
\begin{enumerate}
	\item As CIs are implied by only missing edges, we can not test if an
		incorrect edge is present in the model.
	\item For larger graphs, the number of CI tests can be very large, for example,
		the alarm network with $ 37 $ nodes has $ 287 $ implied CIs. As
		there is no way to rank these CIs, it can be tedious to
		consider all these statements.
\end{enumerate}

% As it is evident that researchers also draw DAGs by hand and even if they use
% an algorithm, the need to make some post-hoc adjustments to the model, in this
% paper we present a tool to assist researchers in making these adjustments
% somewhat inspired by modification indices. In the case of DAGs, global testing
% methodology does exist such as Shipley test, that can combine local structure
% tests into a single p-value test. However, this does not give us a way to make
% modifications to our model. A possible way to do this by checking which of
% these local tests are violated. However, this approach can quickly become
% diffult as the size of the network increases. For example, the popular alarm
% network implies $x$ condition independence statements.

On the other hand, in Structural Equation Models (SEMs) literature, more
emphasis has been placed on helping researchers manually construct models while
integrating their expertise. Researchers typically develop an initial model
based on domain knowledge and use various tools to guide modifications to
improve the fit of their model to data while it still making theoretical sense.
This model construction process is known as \emph{Specification Search}. A
common method for specification search is \emph{modification indices}, which
identify potential changes to the model structure that improve the model fit
and ranks them using the most improvement to the model fit. Researchers then
use these suggestions and their expertise to choose the most optimal
modifications.

\todo[inline]{Show a figure of how the output looks like for modification indices}
\todo[inline]{Other methods: Modification indices and Lagrange Multiplier tests to
selectively add edges And using Wald test or z statistics (also known as
critical ratios) for removing edges.}

@@

Inspired by this ranking approach, in this paper we present a way to rank the CI 
testing using a measure of association. This can be interpreted as the edge

@@

Our main contributions in this papers are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this association measure, we present a procedure to suggest
		modifications (i.e., new edges, and deleting edges) to a given
		model such that the model can better explain the observed
		associations in the data. 
	\item Lastly, we present a web-browser based tool to allow researchers
		to apply this method to their own datasets.
\end{enumerate}

@@

				\draw[->]  (age) to[bend right=20] (ed);
				\draw[->]  (sex) to (ed);
				\draw[->]  (age) to (hrpw);
				\draw[->]  (ed) to (hrpw);
				\draw[->]  (sex) to (hrpw);
				\draw[->]  (ed) to (income);
				\draw[->]  (hrpw) to (income);
				\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
				\draw[->]  (sex) to (income);
				\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
				\draw[->]  (sex) to (occ);	
			\end{scope}
			\begin{scope}[xshift=-3cm]
				\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
				\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
					\begin{minipage}{.2\textwidth}
						\begin{equation*}
							\begin{split}
								\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
								\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
								\textnormal{Age} &\ci \textnormal{Sex} \\
								\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\end{split}
						\end{equation*}
					\end{minipage}
					};
			\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Show a full example of testing using this model}}
		\label{fig:ci_testing}
\end{figure}

@@

\section{Background and Related Work}
\label{sec:background}
We consider random variable $ X $ and $ Y $ with a possible multi-dimensional
variable $ \bm{Z} = \{ Z_1, \cdots, Z_k \} $. We consider these variables in
the mixed data setting where these variables can be any combination of
continuous, categorical, or ordinal unless specified. We use $ \rvert \bm{Z}
\rvert $ to represent the cardinality of the variable $ \bm{Z} $. We write $
\bm{x} = (x_1, \cdots, x_n) $ for a sample from $ X $ of size $ n $. We write
the expectation of a variable $ X $ as $ \mathbb{E}[X] $, conditional
expectations as $ \mathbb{E}[X | \bm{Z}] $.

We consider the problem of estimating the partial (or conditional) association
between $ X $ and $ Y $ in the presence of a set of conditioning variables $
\bm{Z} $. When $ Z = \emptyset $, this is equivalent to the marginal
association between $ X $ and $ Y $.

@@

In this section, we develop a novel partial measure of association for
mixed data by extending the idea of partial correlation to mixed data. We do
this by combining a mixed-data residualization method with a multivariate
measure of association. Given two variables $ X $ and $ Y $, and a set of
conditional variables $ \bm{Z} $, we are interested in computing an estimate of
the conditional association, $ \rho(X, Y; \bm{Z}) $. If $ \bm{Z} = \emptyset $, we
use a constant value as $ Z $. Firstly, we compute the residuals for $ \bm{x} $ and
$ y $ using the $ \bm{z} $ as the predictor variables. We use the mixed data
residualization approach from \citet{Ankan2023}. We consider each variable type
separately and repeat the following steps for both variables $ X $ and $ Y $:

@@

Using this mixed data partial association measure, we propose an iterative DAG
modification / construction procedure. Users can start with an empty graph.
Given a DAG $ G $ and a dataset $ D $, for each pair for variable $ X $ and $ Y
$ in $ G $ that does not have an edge between them, we compute the partial
association measure: $ \rho(X, Y; pa_G(X) \cup pa_G(Y)) $. Additionally, for
each pair of variable in $ G $, that has an edge between them in we compute the
partial association measure: $ \rho(X, Y; pa_{\underline{G}}(X) \cup
pa_{\underline{G}}(Y) $. In the case of no edge, this measure of association
can be interpreted as the unexplained

@@

\begin{theorem}
If we have access to an oracle that can give correct edge directions, the
algorithm above is consistent.
\end{theorem}
